---
title: "NYPD Data Visualization"
output:
  learnr::tutorial:
allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(shiny)
library(readr)
library(plyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(reshape2)
library(scales)
library(ggthemes)


FIG_HEIGHT = 2

####### Read in the nypd dataset #######
nypd_arr <- read_csv("https://raw.githubusercontent.com/skuiper/NYPD-Labs_2020/master/Data/nypd_arrestBar.csv")

#Remove an outlier
nypd_arr <- subset(nypd_arr, pct<=200)

nypd_data <- nypd_arr
nypd_data <- mutate(nypd_data, CrimeType = replace(CrimeType, CrimeType == "Tresspass" ,"Trespass"))
```


<!-- START_SECTION -->
## 1. Getting Started
  
This online tutorial is designed to demonstrate how to use R to create graphs with data from the  New York City Police Department (NYPD). In this page, you will be able to type code into any of the boxes below. If it doesn’t work, simply try a “hint” or click the “start Over” button. If you have never written R code before, you might choose to work through an Introduction to R before attempting this activity.

### Packages and Datasets

This tutorial will focus on the R package ggplot2, created by Hadley Wickham. The key function used in `ggplot2` is `ggplot()` (the *grammar of graphics plot*). This function is different from other graphics functions because it uses a particular grammar inspired by Leland Wilkinson’s landmark book, *The Grammar of Graphics*, which focuses on thinking about, reasoning with, and communicating with graphics. 

Before you start, you will need to install the following packages to complete this tutorial:
  
```{r load_packages, exercise = TRUE, eval = FALSE}
# install.packages("ggplot2")
library(ggplot2)
library(dplyr)
```

**Data**: In this tutorial, we will use the stop-and-frisk reports data from the  New York City Police Department (NYPD). 

The data set is already imported into this tutorial as `nypd_data`.
  
```{r load_datasets, exercise = TRUE, eval = FALSE}
nypd_data 
```
  

To start, let's learn about some variables from the NYPD data:

### Variables of Interest:

* `Gender`: the gender of the individuals that were stopped by NYPD from 2006 through 2014
* `Race` : the race of individuals that were stopped by NYPD from 2006 through 2014.
* `Year`: a selected year to review from the data collected about individuals stopped by NYPD during 2006 through 2014.
* `CrimeType`: type of crime
* `Arrested`: the total number of people arrested by NYPD during the years 2006
through 2014
* `Frisked`: the total number of individuals that were stopped by NYPD and then frisked
from 2006 through 2014
* `Searched`: the total number of individuals that were stopped by NYPD and then
searched from 2006 through 2014
* `Stopped`: the total number of stops made by NYPD from 2006 through 2014.

(The following variables all refer to the types of force that were used to stop or arrest an individual.)

* `HandCuff`	
* `PepperSpray`	
* `Firearm`	
* `CEW`	
* `Verbal`	
* `Summons`
* `Other`	

Run the following code to get a peak of the NYPD data:
  
#### Example 1.1 - loading data
```{r display_data, exercise = TRUE, eval = FALSE}
# The code below uses the head() function to view the first few lines of the NYPD data.
head(nypd_data)
```

```{r display_summary_data, exercise = TRUE, eval = FALSE}
# The code below uses the table() command to see summary data for NYPD data.
table(nypd_data)
```


**Questions:**  

1.1 Try modifying the above code to determine the number of rows in the NYPD data set. 

```{r 11,exercise=TRUE,exercise.lines=3}

```

```{r 11-hint-1}
# Notice that the str() command shows the dimensions of a data set.
```

```{r 11-hint-2}
str(nypd_data)
```


1.2 What code could you use to find the average number of stops from the NYPD data?

```{r 12,exercise=TRUE,exercise.lines=3}

```


```{r 12-hint-1}
# Try the favstats() command with variable Stopped from nypd_data
```

```{r 12-hint-2}
favstats(nypd_data$Stopped)
```

<!-- END_SECTION -->

  
<!-- START_SECTION -->
## 2. Our First Graph

If you worked through the previous lab, you have already used an app to create some bar charts of the NYPD data.

```{r display_plot, echo = FALSE}
ggplot(data = nypd_data, aes(x = Race, y = Arrested)) +
  geom_bar(stat = "identity") +
  labs(title="Number of Arrested by Race")
```

Whenever we are coding it in R it is important to be able to clearly understand the following two questions. 

1. What do you want the computer to do?
2. What must the computer know in order to do that?

To make the above plot, the answers to our questions are

1. **What we want the computer to do:**

      *	Make a bar chart

2. **What the computer needs to know:** The data used for the plot
      
      *	The variable to be plotted along the x-axis.
      *	The variable to be plotted along the y-axis.
      *	The data set that contains the variables.

### Plot Template

All of our graphs will have the same basic template to identify the needed information, as shown below:

<h4 style="text-align: center;">
  ggplot( aes(
  <span style="border: 2px solid gray; padding: 2px;">x</span>
  ,
  <span style="border: 2px solid gray; padding: 2px;">y</span>
  ), data =
  <span style="border: 2px solid gray; padding: 2px;">mydata</span>
  ) + 
  <span style="border: 2px solid gray; padding: 2px;">geom_bar(stat = "identity")</span>
</h4>

<br>
where the pieces in the boxes can be replaced with specific variables to achieve different purposes:

<!-- make the table -->
<div style="display: flex; justify-content: center; align-items: center; margin-bottom: 20px;">
  <table border="1" style="width: 60%; border-collapse: collapse; text-align: center;">
    <tr>
      <th>Box</th>
      <th>Specific Information</th>
      <th>Purpose</th>
    </tr>
    <tr>
      <td>Y</td>
      <td>Stopped</td>
      <td>y-axis variable</td>
    </tr>
    <tr>
      <td>X</td>
      <td>Race</td>
      <td>x-axis variable</td>
    </tr>
    <tr>
      <td>mydata</td>
      <td>nypd_data</td>
      <td>name of data set</td>
    </tr>
    <tr>
      <td>goal</td>
      <td>geom_bar(stat = "identity")</td>
      <td>plot bars</td>
    </tr>
  </table>
</div>


Now, we can put each piece in its place in the template and then run the code to create the plot.

### Make your first graph

#### Example 2.1 - the bar chart of the number of stops by race
```{r create_plot1, exercise = TRUE, eval = FALSE}
# The code below uses ggplot() and geom_bar() to create a bar chart of stops by race
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity")
```

**Questions:**  

2.1 Create a new bar chart showing the count of stops by gender

```{r exercise21, exercise = TRUE}
# Write your code here

```

```{r exercise21-hint}
# Follow the pattern of the above example, changing the variables.
```


<!-- END_SECTION -->

<!-- START_SECTION -->
## 3. Labels and Annotations

Adding labels and titles helps improve the readability of your plots. Let's add some to our bar chart.

#### Example 3.1 - adding your own labels for the x-axis and y-axis of the plot
```{r add_labels1, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adds labels to the bar chart
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity") +
  labs(x = "Race", y = "Number of Stops")
```

Note that the labels of x-aixs and y-axis are set to the name of the corresponding by default, so even if we didn't add our own labels, there are still default labeling.

#### Example 3.2 - adding a title to the plot
```{r add_labels2, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adds a title to the bar chart
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Stops by Race")
```

#### Example 3.3 - combining labels and title
```{r add_labels, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below3 adds labels and a title to the bar chart
ggplot(data = nypd_data, aes(x = Gender, y = Stopped)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Stops by Gender", x = "Gender", y = "Number of Stops")
```



**Questions:**  

3.1 Create a new bar chart showing the count of arrests by year, and add appropriate labels and a title.

```{r exercise31, exercise = TRUE, eval = FALSE}
# Write your code here

```

```{r exercise31-hint-1}
# Follow the same pattern as Example 1 and change the y variable to `Arrested`.
```

```{r exercise31-hint-2}
# Use the `labs()` function to add labels and a title.
```
<!-- END_SECTION -->
  
<!-- START_SECTION -->
## 4. Coloring, Faceting, and Other Aesthetics

After building the basics of the plot, we can use different aesthetics in ggplot2 to enhance the visualizations. For example, we can use color to distinguish data, facet data to create subplots, and adjust the transparency, and type of bars.

### Coloring

The following example shows how coloring helps distinguish between different categories within our data.

#### Example 4.1.1 - coloring the data by gender
```{r example_color1, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below uses color to distinguish data by another variable (e.g., Gender).
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Number of Stops by Race and Gender", x = "Race", y = "Number of Stops")
```


After adding the color aesthetics on the base layer, we can override aesthetics locally within other layers, so that some layers can have different aesthetics than others.

#### Example 4.1.2 - a local override of color
```{r example_color2a, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adds a specific color to the bar chart that overrides the fill property
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "stack", color = "lightblue") +
  labs(title = "Number of Stops by Race and Gender", x = "Race", y = "Number of Stops")
```

```{r example_color2b, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("pink", "yellow4", "cyan")) +
  labs(title = "Number of Stops by Race and Gender", x = "Race", y = "Number of Stops")
```

### Faceting

Faceting allows us to split our plot into multiple subplots based on the levels of some variable in the data set. In the `ggplot2` package, the function `facet_wrap()` is commonly used for faceting the plot.

#### Example 4.2.1 - faceting by year
```{r example_facet1, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below uses `facet_wrap()` to split the plot based on Year.
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Year) +
  labs(title = "Number of Stops by Race and Year", x = "Race", y = "Number of Stops")
```

#### Example 4.2.2 - faceting by gender
```{r example_facet2, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below uses `facet_wrap()` to split the plot based on Year.
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Gender) +
  labs(title = "Number of Stops by Race and Gender", x = "Race", y = "Number of Stops")
```

### Other Aesthetics
Now, let's look at some other ways to adjust the aesthetics of our plot.

#### Example 4.3.1 - adjusting transparency of bars

We can adjust the transparency of the bars using the `alpha` parameter. This can be useful when you want to visualize overlapping bars or simply to make the plot look less dense.

```{r example_aesthetics1, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adjusts the transparency of the bars
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  labs(title = "Number of Stops by Race and Gender with Transparency", x = "Race", y = "Number of Stops")
```

#### Example 4.3.2 - adjusting bar width

The width of the bars can be adjusted using the `width` parameter. This can help in visualizing plots where bar spacing is crucial for clarity.

```{r example_aesthetics2, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adjusts the width of the bars
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.5) +
  labs(title = "Number of Stops by Race and Gender with Adjusted Bar Width", x = "Race", y = "Number of Stops")
```

#### Example 4.3.3 - adding bar border color

Adding a border color to the bars can enhance the visual separation between different categories and make the plot more visually appealing.

```{r example_aesthetics3, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adds a border color to the bars
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  labs(title = "Number of Stops by Race and Gender with Border Color", x = "Race", y = "Number of Stops")
```

**Questions:**  

4.1 Create a new bar chart showing the number of stops by race, faceted by year, and use color to distinguish by gender.

```{r exercise41, exercise = TRUE, eval = FALSE}
# Write your code here


```

```{r exercise41-hint-1}
# Use `facet_wrap(~Year)` to create facets.
```


4.2 Use color to distinguish your plot from 3.1 by gender.

```{r exercise42, exercise = TRUE, eval = FALSE}
# Write your code here


```

```{r exercise42-hint-1}
# Use `fill = Gender` to distinguish by gender.
```

<!-- END_SECTION -->
  
<!-- START_SECTION -->
## 5. Themes
  
Themes are pre-built style templates used to better tailor a graphic to the mode of publication.

### Applying Pre-built Theme

```{r example_theme_1, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below applies the bw theme to the plot
ggplot(data = nypd_data, aes(x = Race, y = Stopped)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(title = "Number of Stops by Race", x = "Race", y = "Number of Stops")
```

Other pre-built themes:
  
* `theme_bw()`
* `theme_linedraw()`, `theme_light()`, and `theme_dark()`
* `theme_minimal()`
* `theme_classic()`
* `theme_void()`

You can judge the differences in these themes below:

```{r, echo = FALSE, fig.height = 4}
library(patchwork)
x <- c("A", "B", "C")
y <- c(1, 2, 3)
data <- data.frame(x, y)

plot1 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_grey()") +
           theme_grey() +
           theme(plot.title = element_text(hjust = 0.5))

plot2 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_bw()") +
           theme_bw() +
           theme(plot.title = element_text(hjust = 0.5))

plot3 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_linedraw()") +
           theme_linedraw() +
           theme(plot.title = element_text(hjust = 0.5))

plot4 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_light()") +
           theme_light() +
           theme(plot.title = element_text(hjust = 0.5))

plot5 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_dark()") +
           theme_dark() +
           theme(plot.title = element_text(hjust = 0.5))

plot6 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_minimal()") +
           theme_minimal() +
           theme(plot.title = element_text(hjust = 0.5))

plot7 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_classic()") +
           theme_classic() +
           theme(plot.title = element_text(hjust = 0.5))

plot8 <- ggplot(data = data, aes(x = x, y = y)) +
           geom_bar(stat = "identity") +
           labs(title = "theme_void()") +
           theme_void() +
           theme(plot.title = element_text(hjust = 0.5))

(plot1 | plot2 | plot3 | plot4) / (plot5 | plot6 | plot7 | plot8)
```

### Customizing Themes

After applying the pre-built theme, we can also add customized features using the `theme()` function. Most commonly this function is used to remove a graph's legend:

#### Example 5.1: removing graph's legend from pre-built theme
```{r example_theme_3, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below adds a bw theme and uses `theme()` to remove the graph's legend
ggplot(data = nypd_data, mapping = aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity") +
 

 theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Number of Stops by Race without Legend", x = "Race", y = "Number of Stops")
```

#### Example 5.2: adding customized features to the theme
```{r example_theme_4, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# The code below customizes font size, color, and type for title and axes
ggplot(data = nypd_data, mapping = aes(x = Race, y = Stopped, fill = Gender)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Stops by Race", x = "Race", y = "Number of Stops") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "gray"),
    axis.title.x = element_text(size = 14, face = "italic", color = "darkblue"),
    axis.title.y = element_text(size = 14, face = "italic", color = "darkblue"),
    axis.text = element_text(size = 12)
  )
```

**Questions:**  

5.1 Use NYPD data to create a bar chart of stops by race and apply the classic theme.

```{r exercise51, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# Write your code here

```

```{r exercise51-hint}
# Use `theme_classic()` to apply the classic theme.
```

5.2 Add 3 customized features to the classic theme

```{r exercise52, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# Write your code here

```

```{r exercise52-hint}
# Use `theme()` to customize features such as `plot.title`, `axis.title.x`, and `axis.title.y`.
```

5.3 Create a bar chart with your choice of theme and add appropriate labels and a title.

```{r exercise53, exercise = TRUE, eval = FALSE}
# Write your code here

```

```{r exercise53-hint-1}
# Choose a theme from the provided list (e.g., `theme_minimal()`).
```

```{r exercise53-hint-2}
# Use `labs()` to add labels and a title.
```

<!-- END_SECTION -->
  
<!-- START_SECTION -->
## 6. Let’s Test Your Skills!

Here are an app that can help you do the data wrangling and create visualizations using the NYPD data. Feel free to play around with them to generate different plots you like and observe the code behind the plots!

(Need the link to embed NYPD shiny app here)
```{r, echo=FALSE}
# knitr::include_app("https://shiny.grinnell.edu/NYPDRTutorial/", height = "500px")
```

**Questions:**  

Try to create the following plots on your own and check your answers using the above apps.

6.1 Create a bar chart showing the number of stops by race, faceted by gender, and customize the theme.

```{r echo =FALSE}
# Expected plot code:
ggplot(data = nypd_data, aes(x = CrimeType, y = Stopped)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~Race) +
  labs(title = "Number of Stops by Crime Type and Race", x = "Crime Type", y = "Number of Stops") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "gray"),
    axis.title.x = element_text(size = 14, face = "italic", color = "darkblue"),
    axis.title.y = element_text(size = 14, face = "italic", color = "darkblue"),
    axis.text = element_text(size = 12)
  )
```

```{r exercise61, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# Write your code here

```

```{r exercise61-hint-1}
# Use `facet_wrap(~Gender)` to create facets by gender.
```

```{r exercise61-hint-2}
# Customize the theme using `theme_minimal()` and additional `theme()` customizations.
```

6.2 Create a bar chart showing the number of stops by race, colored by year, with customized bar width and transparency.

```{r echo =FALSE}
# Expected plot code:
ggplot(data = nypd_data, aes(x = Race, y = Stopped, fill = factor(Year))) +
  geom_bar(stat = "identity", position = "stack", width = 0.7, alpha = 0.8) +
  labs(title = "Number of Stops by Race and Year", x = "Race", y = "Number of Stops") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "black"),
    axis.title.x = element_text(size = 14, face = "italic", color = "black"),
    axis.title.y = element_text(size = 14, face = "italic", color = "black"),
    axis.text = element_text(size = 12)
  )
```


```{r exercise62, exercise = TRUE, eval = FALSE, fig.height = FIG_HEIGHT}
# Write your code here

```

```{r exercise62-hint-1}
# Use `fill = factor(Year)` to color by year.
```

```{r exercise62-hint-2}
# Adjust `width` and `alpha` in `geom_bar()` for customized bar width and transparency.
```


By completing these exercises, you will practice creating bar charts, customizing themes, and using faceting and coloring to enhance your visualizations.
<!-- END_SECTION -->
